How environmental drivers of spatial synchrony interact

### 2023-06-27 Paper/Paper.Rnw
various changes

Data for the project are publicly archived \citep{SBCLTER2021}.
All computations were done in R version 3.6.3 on a laptop running Ubuntu Linux
16.04. Complete codes for the project workflow will be published online, simultaneously with the final version of the paper.
%are at \url{https://github.com/reumandc/InteractingMoranEffects.git}.

@@

%Fourier versus wavelet
Since both our Fourier approach and the wavelet methods of \cite{CastoraniSheppardKelp} provided
similar results when applied to kelp, when should one approach be applied in favor of the other?
The Fourier approach of our study was designed to facilitate 
mathematical examination of interactions between Moran effect as a general mechanism, and development
of general theory. The Fourier
approach facilitates deriving analytic results from mathematical models such as (\ref{main_model_4}).
The wavelet approach of \cite{CastoraniSheppardKelp} was instead optimized for detecting interactions 
and identifying Moran mechanisms in data, in spite of non-stationarity and
other complicating features which are present in many ecological datasets. 
The study of
\cite{Sheppard2019} developed the wavelet methods applied by \cite{CastoraniSheppardKelp}; an open-source implementation of 
these methods \citep{wsyn} can facilitate future work.
The modelling approach of this study relates indirectly to the approach of \cite{Anderson2021}, though that
study concerned different research questions.

@@

%nonlinearity
Our model is a linear population dynamical model, but most population dynamics are probably 
nonlinear, at least to some degree. We speculate on what may be the outcome if future work
explores interactions between two or more Moran effects using nonlinear population dynamical 
models. 
Some researchers have explored Moran effects of one environmental variable acting on nonlinear
population models and found generally that synchronous environmental fluctuations tend to promote
synchronous population fluctuations, in spite of nonlinearities, though often to a lesser 
degree than for the linear case \citep{Royama2005,Engen2005}. 
Our model is a linearization of a very general population dynamical model [SI section
S1.2 of \cite{Desharnais2018} and SI section S1 of \cite{Walter2017}], and it is well known
that analytic results for linearized models, such as we have developed, also hold, approximately, for
the original nonlinear model, as long as the noise impinging the model is sufficiently weak. 
How weak is ``sufficiently weak'' is model dependent, but prior work examining that question
provided evidence that weak noise approximations using linearizations of commonly used 
ecological models are often actually quite accurate, even for fairly strong 
noise \cite{Nisbet1977}. 
Therefore we speculate that our main observations will hold even for nonlinear systems:
1) Moran effects can substantially interact, either synergistically or destructively;
2) interactions can depend on timescale; and 3) the nature of effects depend on relative,
timescale-specific lags of environmental influences on populations and between 
environmental variables. However, future work should nevertheless examine Moran interactions and 
nonlinear models, since much has been learned by exploring deviations of ecological models 
from predictions of theory based on linearization (e.g., \citealp{Reuman2006}).

@@

\section{Author contributions}
DCR conceptualized the paper, developed methodology, performed formal analysis and 
investigation, carried out project administration and supervision, provided computing 
resources, wrote and validated the software, produced most of the figures and wrote the 
first draft. KCC and TWB performed data curation. MCNC helped with data visualization 
and curation and project conceptualization. JAW and LWS helped with conceptualization, 
methodology and formal analysis. MCNC and JAW helped with project administration. 
All authors acquired the funding and performed substantial editing of manuscript drafts.

>>>>

Data for the project are publicly archived \citep{SBCLTER2021}.
All computations were done in R version 3.6.3 on a laptop running Ubuntu Linux
16.04. Complete codes for the project workflow are at \url{https://github.com/reumandc/InteractingMoranEffects.git}.

@@

%Fourier versus wavelet
Since both our Fourier approach and the wavelet methods of \cite{CastoraniSheppardKelp} provided
similar results when applied to kelp, when should one approach be applied in favor of the other?
The Fourier approach of our study was designed to facilitate 
mathematical examination of interactions between Moran effects as a general mechanism, and development
of general theory. The Fourier
approach facilitates deriving analytic results from mathematical models such as (\ref{main_model_4}).
The wavelet approach of \cite{CastoraniSheppardKelp} was instead optimized for detecting interactions 
and identifying Moran mechanisms in data, in spite of non-stationarity and
other complicating features which are present in many ecological datasets. 
The study of
\cite{Sheppard2019} developed the wavelet methods applied by \cite{CastoraniSheppardKelp}; an open-source implementation of 
these methods \citep{wsyn} can facilitate future work.
The modelling approach of this study relates indirectly to the approach of \cite{Anderson2021}, though that
study concerned different research questions.

@@

%nonlinearity
Our model is a linear population dynamical model, but most population dynamics are probably 
nonlinear, at least to some degree. We speculate on what may be the outcome if future work
explores interactions between two or more Moran effects using nonlinear population dynamical 
models. 
Some researchers have explored Moran effects of one environmental variable acting on nonlinear
population models and found generally that synchronous environmental fluctuations tend to promote
synchronous population fluctuations, in spite of nonlinearities, though often to a lesser 
degree than for the linear case \citep{Royama2005,Engen2005}. 
Our model is a linearization of a very general population dynamical model [SI section
S1.2 of \cite{Desharnais2018} and SI section S1 of \cite{Walter2017}], and it is well known
that analytic results for linearized models, such as we have developed, also hold, approximately, for
the original nonlinear model, as long as the noise impinging the model is sufficiently weak. 
How weak is ``sufficiently weak'' is model dependent, but prior work examining that question
provided evidence that weak noise approximations using linearizations of commonly used 
ecological models are often actually quite accurate, even for fairly strong 
noise \citep{Nisbet1977}. 
Therefore we speculate that our main observations will hold even for nonlinear systems:
1) Moran effects can substantially interact, either synergistically or destructively;
2) interactions can depend on timescale; and 3) the nature of effects depend on relative,
timescale-specific lags of environmental influences on populations and between 
environmental variables. However, future work should nevertheless examine Moran interactions and 
nonlinear models, since much has been learned by exploring deviations of ecological models 
from predictions of theory based on linearization (e.g., \citealp{Reuman2006}).

@@

\section{Author contributions}
\textbf{Reuman:} Conceptualization (lead), formal analysis (lead), funding acquisition (equal), investigation (lead), methodology (lead), project administration (equal), resources (lead), software (lead), validation (lead), visualization (lead), writing - original draft (lead), writing - review and editing (lead). 
\textbf{Castorani:} Conceptualization, data curation, funding acquisition (equal), project administration (equal), visualization, writing - original draft, writing - review and editing.
\textbf{Cavanaugh:} Data curation, funding acquisition (equal), writing - review and editing. 
\textbf{Sheppard:} Conceptualization, formal analysis, funding acquisition (equal), methodology, writing - review and editing.
\textbf{Walter:} Conceptualization, formal analysis, funding acquisition (equal), methodology, project administration (equal), writing - review and editing.
\textbf{Bell:} Data curation (lead), funding acquisition (equal), writing - review and editing.
